---
title: "Ordinal 2.0"
author: "Andrés Felipe Tigreros Andrade"
format: html
editor: visual
---

## When Eavesdropping Fades: Spatial Degradation of Alarm Signals in Amazonian Mixed-species Flocks

Here, we are going to analyze the data obtained from experiments explained in the `README.md` document. This is a different quarto in which I will analyze the data based on Daniel T. Blumstein recommendations after our meeting on 02/April/2026.

## 1. Starting with the setup

### 1.1. Activating packages

We will activate the dataset and establish the working locationisables the printing of code (only output is displayed).

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(patchwork)
library(lme4)
library(Matrix)
library(betareg)
library(glmmTMB)
library(AICcmodavg)
library(lmerTest)
library(extrafont)
library(lmtest)
library(fastDummies)
library(janitor)
library(mclogit)
library(scales)
library(here)
library(brms)
library(ggeffects)
library(lme4)
library(ordinal)
library(MASS)
```

### 1.2. Reading the dataset

We are going to read the dataset in which we filled all of the data collected from the experiments. We will call `janitor::clean_names()` in order to have decent column names. Then, use `mutate()` in order to create a column named "real_d" in which we sum both distances to have the actual distance measured in the field. The last use of mutate will be to create or assign a name to each of the levels of the response designed for *M. longipennis.*

```{r}
data <- read_csv(file = "finalized_playbacks.csv") %>% 
  janitor::clean_names() %>%  ##making better column names
  mutate(real_d = distance1 + distance2,  ##creating a column with real distance
         response_cat = factor(rpta_true, levels = c(0, 1, 2, 3, 4), 
                               labels = c("No Response", "Dive", "Freeze", "Look Around", "Countercall")))          
  
head(data)
str(data)
```
