---
title: "Ordinal 2.0"
author: "Andrés Felipe Tigreros Andrade"
format: html
editor: visual
---

## When Eavesdropping Fades: Spatial Degradation of Alarm Signals in Amazonian Mixed-species Flocks

Here, we are going to analyze the data obtained from experiments explained in the `README.md` document. This is a different quarto in which I will analyze the data based on Daniel T. Blumstein recommendations after our meeting on 02/April/2026.

## 1. Starting with the setup

### 1.1. Activating packages

We will activate the dataset and establish the working locationisables the printing of code (only output is displayed).

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(patchwork)
library(lme4)
library(Matrix)
library(betareg)
library(glmmTMB)
library(AICcmodavg)
library(lmerTest)
library(extrafont)
library(lmtest)
library(fastDummies)
library(janitor)
library(mclogit)
library(scales)
library(here)
library(brms)
library(ggeffects)
library(lme4)
library(ordinal)
library(MASS)
```

### 1.2. Reading the dataset and tidying it

We are going to read the dataset in which we filled all of the data collected from the experiments. We will call `janitor::clean_names()` in order to have decent column names. Then, use `mutate()` in order to create a column named "real_d" in which we sum both distances to have the actual distance measured in the field. The last use of mutate will be to create or assign a name to each of the levels of the response designed for *M. longipennis.*

```{r}
data <- read_csv(file = "finalized_playbacks.csv") %>% 
  janitor::clean_names() %>%  ##making better column names
  mutate(real_d = distance1 + distance2,  ##creating a column with real distance
         response_cat = factor(rpta_true, levels = c(0, 1, 2, 3, 4), 
                               labels = c("No Response", "Dive", "Freeze", "Look Around", "Countercall")))          
  
head(data)
str(data)
skimr::skim(data)
```

Now, I will drop off the controls ("CONTRL") and the Counter Call response due to not being that common.

```{r}
data <- data %>% filter(pb_spp != "CONTRL") %>% 
  filter(rpta_true != "4")
```

### 1.3. Adding the order of the levels

For this part of the code, the purpose is to assign the order of the factors. The **A** will be maintaining all of the levels from the 0: no response; 1: look around (considered less intense because is not an actual evasive behavior); 2: freeze (considered as a maneuver not as intense because the individual doesn't feel threatened at is full) and 3: dive (considered as the most frenetic maneuver). The **B** is Ari's suggestion. In his suggestion, he considered that both Freeze and Dive can be considered as responses with the same intensity that will depends on the context. If you're far away from a shelter, you will freeze so the predator will not notice your movements. If you're close to the floor or a shelter, you will dive down as escaping from the predator.

```{r}
data1 <- data %>% mutate(
  # Ordering A: 4-level ordinal by commitment (My assessment)
  response_a = factor(
    rpta_true,
    levels  = c(0, 3, 2, 1),
    labels  = c("no_response", "look_around", "freeze", "dive"),
    ordered = TRUE
  ),
  # Ordering B: 3-level (freeze and dive merged) (Ari's assessment)
  response_b = factor(
    case_when(
      rpta_true == 0           ~ "no_response",
      rpta_true == 3           ~ "look_around",
      rpta_true %in% c(1, 2)   ~ "evasive"
    ),
    levels  = c("no_response", "look_around", "evasive"),
    ordered = TRUE
  ),
  distance_c = real_d - 25,         # center distance at midpoint
  pb_spp     = factor(pb_spp),
  flock      = factor(flock_name)
)

ggplot(data = data1, aes(x = distance_c)) +
  geom_histogram()

str(data1)
```

Sanity checks

```{r}
data1 %>%
  count(pb_spp, real_d, response_a) %>%
  pivot_wider(names_from = response_a, values_from = n, values_fill = 0) %>%
  print(n = Inf)

any(is.na(data1$response_a))
any(is.na(data1$response_b))
any(is.na(data1$real_d))
any(is.na(data1$pb_spp))

# trials per flock
data1 %>% count(flock_name) %>% 
  arrange(desc(n))
```

## 2. Some graphs

```{r}
# Ordering A
p_compA <- df_clean %>%
  mutate(distance = factor(distance, levels = c(10, 20, 30, 40))) %>%
  count(pb_spp, distance, response_a) %>%
  group_by(pb_spp, distance) %>%
  mutate(prop = n / sum(n),
         total = sum(n)) %>%
  ungroup() %>%
  ggplot(aes(x = distance, y = prop, fill = response_a)) +
  geom_col(position = "stack", width = 0.75, color = "white", linewidth = 0.3) +
  geom_text(aes(label = paste0("n=", total), y = 1.04),
            data = . %>% distinct(pb_spp, distance, total),
            size = 3, color = "grey30") +
  facet_wrap(~ pb_spp) +
  scale_fill_brewer(palette = "YlOrRd", direction = -1,
                    labels = c("No response", "Look around", "Freeze", "Dive")) +
  scale_y_continuous(labels = scales::percent_format(), expand = c(0, 0),
                     limits = c(0, 1.08)) +
  labs(x = "Distance (m)", y = "Proportion of trials",
       fill = "Response",
       title = "Ordering A: 4-level response") +
  theme_bw(base_size = 11) +
  theme(panel.grid.major.x = element_blank(),
        panel.grid.minor   = element_blank(),
        strip.background   = element_rect(fill = "grey90", color = NA),
        legend.position    = "right")

# Ordering B
p_compB <- df_clean %>%
  mutate(distance = factor(distance, levels = c(10, 20, 30, 40))) %>%
  count(pb_spp, distance, response_b) %>%
  group_by(pb_spp, distance) %>%
  mutate(prop = n / sum(n),
         total = sum(n)) %>%
  ungroup() %>%
  ggplot(aes(x = distance, y = prop, fill = response_b)) +
  geom_col(position = "stack", width = 0.75, color = "white", linewidth = 0.3) +
  geom_text(aes(label = paste0("n=", total), y = 1.04),
            data = . %>% distinct(pb_spp, distance, total),
            size = 3, color = "grey30") +
  facet_wrap(~ pb_spp) +
  scale_fill_brewer(palette = "YlOrRd", direction = -1,
                    labels = c("No response", "Look around", "Evasive")) +
  scale_y_continuous(labels = scales::percent_format(), expand = c(0, 0),
                     limits = c(0, 1.08)) +
  labs(x = "Distance (m)", y = "Proportion of trials",
       fill = "Response",
       title = "Ordering B: 3-level response (freeze + dive collapsed)") +
  theme_bw(base_size = 11) +
  theme(panel.grid.major.x = element_blank(),
        panel.grid.minor   = element_blank(),
        strip.background   = element_rect(fill = "grey90", color = NA),
        legend.position    = "right")

p_compA / p_compB

p_compA
p_compB
p_compA / p_compB

```

## 3. Starting with the models

### 3.1. Frequentist

```{r}
# Ordering A
mA <- clmm(
  response_a ~ distance_c * pb_spp + (1 | flock),
  data = data1,
  link = "logit"
)

# Ordering B
mB <- clmm(
  response_b ~ distance_c * pb_spp + (1 | flock),
  data = data1,
  link = "logit"
)

summary(mA)
summary(mB)

# Optional: simpler models without the interaction, for likelihood-ratio
mA_main <- clmm(response_a ~ distance_c + (1 | flock),
                data = data1, link = "logit")
anova(mA_main, mA)   # is the interaction needed?
```

### 3.1. Bayesian

```{r}
# Ordering A
fit_a <- brm(
  bf(response_a ~ distance_c * pb_spp + (1 | flock),
     disc ~ distance_c),
  data    = data1,
  family  = cumulative("logit"),
  chains  = 4, cores = 4,
  iter    = 4000, warmup = 2000,
  control = list(adapt_delta = 0.95),
  seed    = 42
)

summary(fit_a)
plot(fit_a)        # trace plots — check mixing
pp_check(fit_a)    # posterior predictive check

# Ordering B
fit_b <- brm(
  bf(response_b ~ distance_c * pb_spp + (1 | flock),
     disc ~ distance_c),
  data    = data1,
  family  = cumulative("logit"),
  chains  = 4, cores = 4,
  iter    = 4000, warmup = 2000,
  control = list(adapt_delta = 0.95),
  seed    = 42
)

summary(fit_b)
pp_check(fit_b)
```
